---
title: "QBIO 490 Nick Leon Bruk"
author: "Nick Xu, Leon Chen, Bruk Tefera"
date: "2025-12-06"
output: html_document
---
```{r}
knitr::opts_knit$set(root.dir = normalizePath("/home1/nickxu/490_cluster/analysis_data"))  
```

#I remove the ggsave from the code because everyone has very different directory. It might interfere with running the code when you are grading it :) Nick 

#Nick Xu did OS Clustering, clinical analysis, and MAF analysis. Leon did methylation analysis. Bruk did transcrptomics. Nick combine the code together
```{r}
if (!require("knitr", quietly = TRUE)) 
install.packages("knitr") 
if (!require("BiocManager", quietly = TRUE)) 
install.packages("BiocManager") 
BiocManager::install(version = "3.21") 
if (!require("TCGAbiolinks", quietly = TRUE))  
BiocManager::install("TCGAbiolinks")
if (!require("maftools", quietly = TRUE)) 
BiocManager::install("maftools")
if (!require("sesameData", quietly = TRUE))
BiocManager::install("sesameData")
if (!require("sesame", quietly = TRUE))
BiocManager::install("sesame")
if (!require("limma", quietly = TRUE))
BiocManager::install("limma")
if (!require("dplyr", quietly = TRUE))
BiocManager::install("dplyr")
if (!require("stringr", quietly = TRUE))
BiocManager::install("stringr")
if (!require("survival", quietly = TRUE))
BiocManager::install("survival")
if (!require("survminer", quietly = TRUE))
BiocManager::install("survminer")
if (!require("ggplot2", quietly = TRUE))
BiocManager::install("ggplot2")
if (!require("ggrepel", quietly = TRUE))
BiocManager::install("ggrepel")
if (!require("SummarizedExperiment", quietly = TRUE))
BiocManager::install("SummarizedExperiment")
if (!require("DESeq2", quietly = TRUE))
BiocManager::install("DESeq2")
if (!require("pheatmap", quietly = TRUE))
BiocManager::install("pheatmap")
if (!require("httr", quietly = TRUE))
BiocManager::install("httr")
if (!require("org.Hs.eg.db", quietly = TRUE))
BiocManager::install("org.Hs.eg.db")
if (!require("AnnotationDbi", quietly = TRUE))
BiocManager::install("AnnotationDbi")
if (!require("tibble", quietly = TRUE))
BiocManager::install("tibble")
```

```{r}
library(knitr)
library(BiocManager)
library(TCGAbiolinks)
library(maftools)
library(sesameData)
library(sesame)
library(limma)
library(dplyr)
library(stringr)
library(survival)
library(survminer)
library(ggplot2)
library(ggrepel)
library(SummarizedExperiment)
library(DESeq2)
library(pheatmap)
library(httr)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(tibble)

```

#Clinical Data Downloading
```{r}
clin_query <- GDCquery(project = "TCGA-GBM",                       
                       data.category = "Clinical",                       
                       data.type = "Clinical Supplement",                        
                       data.format = 'BCR Biotab') 
GDCdownload(clin_query)
clinical.BCRtab.all <- GDCprepare(clin_query) 
clinic <- clinical.BCRtab.all$clinical_patient_gbm[-c(1,2),] 
drug   <- clinical.BCRtab.all$clinical_drug_gbm[-c(1,2),]
```

#Drug and patient indication based on TMZ
```{r}
drug$drug_name <- tolower(drug$pharmaceutical_therapy_drug_name)
drug$tmz_status <- grepl("temozolomide", drug$drug_name)

drug <- mutate(
  drug,
  drug_name    = tolower(pharmaceutical_therapy_drug_name),
  tmz_status = grepl("temozolomide", drug_name)
)

```

#Grouping patient based on Drug use and filter TMZ patient 
```{r}
drug_grouped <- group_by(drug, bcr_patient_barcode)

tmz_by_patient <- summarise(drug_grouped, tmz_treated = any(tmz_status, na.rm = TRUE))
```

#Add TMZ column to clinic table. Create TMZ and NO_TMZ factor for future graphing 
```{r}
clinic <- clinic %>%
  left_join(tmz_by_patient, by = "bcr_patient_barcode") %>%
  mutate(
    tmz_treated = ifelse(is.na(tmz_treated), FALSE, tmz_treated),
    tmz_group   = ifelse(tmz_treated, "TMZ", "No_TMZ"),
    tmz_group   = factor(tmz_group, levels = c("No_TMZ", "TMZ"))
  )
```

#Visualization
```{r}
table(clinic$tmz_group)
```

#Compute OS
```{r}
clinic <- clinic %>%
  mutate(
    last_contact_days_to = as.numeric(last_contact_days_to),
    death_days_to = as.numeric(death_days_to),
    OS_event = ifelse(vital_status == "Dead", 1, 0),
    OS_days = ifelse(OS_event == 1, death_days_to , last_contact_days_to),
    OS_months = OS_days / 30.44
  )%>%
  filter(!is.na(OS_months))
```

#Filter only TMZ treated patient
```{r}
clin_tmz <- filter(clinic, tmz_treated == TRUE)
```

#Assign Responder Groups 
```{r}
clin_tmz$responder<- ifelse(
  clin_tmz$OS_months >= 18, "Good_responder",
  ifelse(clin_tmz$OS_months <= 12, "Poor_responder", "Intermediate")
)

clin_tmz_filtered <- clin_tmz %>%
  filter(responder != "Intermediate")

```

```{r}
table(clin_tmz_filtered$responder)
```

#KM plot Good vs Poor 
```{r}
survival_object <- Surv(time = clin_tmz_filtered$OS_months,event = clin_tmz_filtered$OS_event)
fit_object <- survfit(survival_object ~ responder, data = clin_tmz_filtered)

survplot <- ggsurvplot(fit_object,
                       pval=TRUE,
                       ggtheme = theme(plot.margin = unit(c(1,1,1,1), "cm")),
                       legend = 'right')

KM_plot <- survplot$plot + theme_bw() + theme(axis.title = element_text(size=20), 
                                              axis.text = element_text(size=16),
                                              legend.title = element_text(size=14),
                                              legend.text = element_text(size=12))+
  labs(
    x = "Overall Survival (months)",
    y = "Survival Probability",
    title = "Kaplan–Meier: Good vs Poor Responders"
  )

KM_plot




```

#KM plot TMZ vs non_TMZ
```{r}
survival_object <- Surv(time = clinic$OS_months,event = clinic$OS_event)

fit_object <- survfit(survival_object ~ tmz_group, data = clinic)

survplot <- ggsurvplot(fit_object,
                       pval=TRUE,
                       ggtheme = theme(plot.margin = unit(c(1,1,1,1), "cm")),
                       legend = 'right')

KM_plot <- survplot$plot + theme_bw() + theme(axis.title = element_text(size=20), 
                                              axis.text = element_text(size=16),
                                              legend.title = element_text(size=14),
                                              legend.text = element_text(size=12))+

  labs(
    x = "Overall Survival (months)",
    y = "Survival Probability",
    title = "Kaplan–Meier: TMZ vs No TMZ"
  )

KM_plot

```

#MAF Analysis
```{r}
names(clin_tmz_filtered)[names(clin_tmz_filtered) == "bcr_patient_barcode"] <- "Tumor_Sample_Barcode"
```

#Dowload MAf
```{r}
maf_query <- GDCquery(
  project = "TCGA-GBM",
  data.category = "Simple Nucleotide Variation",
  access = "open",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)

GDCdownload(maf_query, method = "api", files.per.chunk = 100)
maf_data <- GDCprepare(maf_query)
maf_object <- read.maf(
  maf = maf_data,
  clinicalData = clin_tmz_filtered,
  isTCGA = TRUE
)
```

#Define Mask
```{r}
mask_good <- maf_object@clinical.data$responder == "Good_responder"
mask_poor <- maf_object@clinical.data$responder == "Poor_responder"
```

#Subset
```{r}
maf_good <- subsetMaf(maf = maf_object, tsb = maf_object@clinical.data$Tumor_Sample_Barcode[mask_good])
maf_poor <- subsetMaf(maf = maf_object, tsb = maf_object@clinical.data$Tumor_Sample_Barcode[mask_poor])
```

#Co-oncoplot
```{r}
coOncoplot(
  m1 = maf_good,
  m2 = maf_poor,
  m1Name = "Good Responders",
  m2Name = "Poor Responders"
)

```

#Co-lollipop
```{r}
par(mfrow = c(3,1))

lollipopPlot2(m1 = maf_good, m2 = maf_poor,
              m1_name = "Good Responders",
              m2_name = "Poor Responders",
              gene = "PTEN")

lollipopPlot2(m1 = maf_good, m2 = maf_poor,
              m1_name = "Good Responders",
              m2_name = "Poor Responders",
              gene = "TP53")

lollipopPlot2(m1 = maf_good, m2 = maf_poor,
              m1_name = "Good Responders",
              m2_name = "Poor Responders",
              gene = "EGFR")

par(mfrow = c(1,1))

```

#Contingency table
```{r}
get_mut_status <- function(maf_object, gene){
  mutated_samples <- subsetMaf(
    maf = maf_object,
    genes = gene,
    fields = "Tumor_Sample_Barcode"
  )@data$Tumor_Sample_Barcode

  clin_tmz_filtered$Tumor_Sample_Barcode %in% mutated_samples
}

clin_tmz_filtered$PTEN_mut <- get_mut_status(maf_object, "PTEN")
clin_tmz_filtered$TP53_mut <- get_mut_status(maf_object, "TP53")
clin_tmz_filtered$EGFR_mut <- get_mut_status(maf_object, "EGFR")

pten_table <- table(clin_tmz_filtered$responder, clin_tmz_filtered$PTEN_mut)
tp53_table <- table(clin_tmz_filtered$responder, clin_tmz_filtered$TP53_mut)
egfr_table <- table(clin_tmz_filtered$responder, clin_tmz_filtered$EGFR_mut)

cat("PTEN Contingency Table:\n"); print(pten_table)
cat("\nTP53 Contingency Table:\n"); print(tp53_table)
cat("\nEGFR Contingency Table:\n"); print(egfr_table)

cat("\nPTEN Fisher Test:\n")
print(fisher.test(pten_table))

cat("\nTP53 Fisher Test:\n")
print(fisher.test(tp53_table))

cat("\nEGFR Fisher Test:\n")
print(fisher.test(egfr_table))

par(mfrow = c(1, 3))  # 3 plots in one row

# PTEN barplot
barplot(
  pten_table,
  beside = TRUE,
  col = c("skyblue", "salmon"),
  legend = TRUE,
  main = "PTEN Mutation vs Responders",
  xlab = "Mutation Status (FALSE = No, TRUE = Yes)",
  ylab = "Patient Count"
)

# TP53 barplot
barplot(
  tp53_table,
  beside = TRUE,
  col = c("skyblue", "salmon"),
  legend = TRUE,
  main = "TP53 Mutation vs Responders",
  xlab = "Mutation Status",
  ylab = "Patient Count"
)

# EGFR barplot
barplot(
  egfr_table,
  beside = TRUE,
  col = c("skyblue", "salmon"),
  legend = TRUE,
  main = "EGFR Mutation vs Responders",
  xlab = "Mutation Status",
  ylab = "Patient Count"
)

par(mfrow = c(1, 1))  # reset plotting layout

```

#Methylation Analysis
```{r}
#Download GBM Methylation Data
library(TCGAbiolinks)

meth_query <- GDCquery(
  project = "TCGA-GBM",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450",
)

GDCdownload(meth_query)

meth_data <- GDCprepare(meth_query)


#check the data
View(meth_data)

# Extract Beta values (CpG sites x samples)
beta_matrix <- assay(meth_data, 1) 

#Can't view it for some reason, check dimensions
dim(beta_matrix)

#Extract sample metadata
sample_info <- colData(meth_data)
head(sample_info$bcr_patient_barcode)
```

```{r}
#Subsetting TMZ patients into good and poor responders
#Extract patient carcodes
sample_info$patient_barcode <- substr(sample_info$bcr_patient_barcode, 1, 12)
head(sample_info$patient_barcode)

# Keep only methylation samples for TMZ-treated patients
tmz_patients <- rownames(clin_tmz_filtered)
meth_tmz_samples <- sample_info$bcr_patient_barcode[sample_info$patient_barcode %in% tmz_patients]

length(meth_tmz_samples)
head(meth_tmz_samples)

# Truncate both methylation matrix and selected samples to first 15 characters
beta_sample_ids <- substr(colnames(beta_matrix), 1, 15)
meth_tmz_samples_short <- substr(meth_tmz_samples, 1, 15)

# Subset beta matrix for TMZ patients
beta_tmz <- beta_matrix[, beta_sample_ids %in% meth_tmz_samples_short]
dim(beta_tmz)
head(colnames(beta_tmz))

#Match good vs poor response to each filtered patient
responder_status <- clin_tmz_filtered$responder[
  match(
    substr(colnames(beta_tmz), 1, 12),  
    rownames(clin_tmz_filtered)
  )
]

#Factorize
responder_status <- factor(responder_status, levels = c("Poor_responder", "Good_responder"))
table(responder_status)

#Transform into beta-values for graphical analysis, as done in the methylation companion
#Filtered NAs
beta_tmz_clean <- beta_tmz[complete.cases(beta_tmz), ]
#Avoid 0/1s
beta_tmz_safe <- pmin(pmax(beta_tmz_clean, 0.001), 0.999)
mval <- t(apply(beta_tmz_safe, 1, function(x) log2(x / (1 - x))))

#Check dimensions and a small preview
dim(mval)
head(mval[, 1:5])
```

```{r}
#Design matrix for Good vs Poor responders
design <- model.matrix(~ responder_status)
colnames(design) <- c("Intercept", "Good_vs_Poor")

#Need to fit the linear model
fit <- lmFit(mval, design)
#Other step from methylation companion
fit <- eBayes(fit)

# Extract top differentially methylated CpGs
top_cpgs <- topTable(fit, coef = "Good_vs_Poor", adjust.method = "BY", number = 50)
#head(top_cpgs)
```

#Volcano plot
```{r}
# Create threshold for highlighting significant CpGs
top_cpgs$threshold <- as.factor(abs(top_cpgs$logFC) < 1)  # TRUE = not significant, FALSE = significant

# Set colors
cols <- c("TRUE" = "grey", "FALSE" = "blue")

# Make the volcano plot
volcano_plot <- ggplot(top_cpgs, aes(x = logFC, y = -log10(adj.P.Val), color = threshold)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_colour_manual(values = cols) +
  geom_vline(xintercept = 1, colour="#990000", linetype="dashed") +
  geom_vline(xintercept = -1, colour="#990000", linetype="dashed") +
  geom_hline(yintercept = 2, colour="#990000", linetype="dashed") +
  theme_bw() +
  theme(legend.position = "none") +
  xlab("M-value Fold Change (Good vs Poor)") +
  ylab("-log10(BY-adjusted p-value)")


print(volcano_plot)
```

#result table of all CpG's
```{r}
full_tt <- topTable(fit, coef = "Good_vs_Poor", number = Inf, adjust.method = "BY")
full_tt$CpG <- rownames(full_tt)

#annotation from sesame?
annot_df <- as.data.frame(rowRanges(meth_data))
annot_df$CpG <- rownames(annot_df)

colnames(annot_df)

annotation_column <- "gene"

full_annotated <- left_join(
  full_tt,
  annot_df[, c("CpG", annotation_column)],
  by = "CpG"
)

#gene name cleaned
full_annotated$geneName <- full_annotated[[annotation_column]]
full_annotated$geneName <- ifelse(
  is.na(full_annotated$geneName) | full_annotated$geneName == "",
  NA_character_,
  full_annotated$geneName
)

#splitting gene strings with multiple genes
full_annotated$geneName <- sapply(
  strsplit(full_annotated$geneName, ";"),
  function(x) if (length(x) >= 1) x[1] else NA_character_
)

#dataframe, from methylation companion assignment
dat <- data.frame(
  foldchange = full_annotated$logFC,
  logPvalue  = -log10(p.adjust(full_annotated$P.Value, method = "BY")),
  geneName   = full_annotated$geneName,
  stringsAsFactors = FALSE
)

dat$threshold <- as.factor(abs(dat$foldchange) < 1)

#plotting
cols <- c("TRUE" = "grey", "FALSE" = "blue")

volcano_plot1 <- ggplot(data = dat, aes(x = foldchange, y = logPvalue, color = threshold)) +
  geom_point(alpha = .2, size = 0.6) +
  scale_colour_manual(values = cols) +
  geom_vline(xintercept = 1, colour = "#990000", linetype = "dashed") +
  geom_vline(xintercept = -1, colour = "#990000", linetype = "dashed") +
  geom_hline(yintercept = 2, colour = "#990000", linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "none") +
  xlab("Fold Change (M-value)") +
  ylab("-log10(BY-adjusted p-value)")

print(volcano_plot1)

#labeled plot
top_hits <- dat[order(-dat$logPvalue), ] %>% 
  head(15)   # choose top 15 significant CpGs

volcano_labeled <- volcano_plot1 +
  geom_text_repel(
    data = top_hits,
    aes(x = foldchange, y = logPvalue, label = geneName),
    size = 3,
    max.overlaps = 50
  )

print(volcano_labeled)
```

#RNASeq Downloading
```{r}
rna_query <- GDCquery(
  project = "TCGA-GBM",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)

GDCdownload(rna_query)
rna_data <- GDCprepare(rna_query)
```

#Transcriptomics Analysis
```{r}
counts_matrix <- assay(rna_data)

sample_ids <- substr(colnames(counts_matrix), 1, 12)
colnames(counts_matrix) <- sample_ids

# Build clinical annotation dataframe for DESeq2: patient barcode + TMZ response group
col_data <- clin_tmz_filtered %>%
  dplyr::select(bcr_patient_barcode, responder) %>%
  dplyr::distinct() %>%
  tibble::column_to_rownames("bcr_patient_barcode")

common_samples <- intersect(colnames(counts_matrix), rownames(col_data))

# Keep only samples appearing in both datasets 
counts_matrix <- counts_matrix[, common_samples]
col_data <- col_data[common_samples, , drop = FALSE]

# Create DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(
  countData = counts_matrix,
  colData   = col_data,
  design    = ~ responder
)

dds <- DESeq(dds)
vsd <- vst(dds, blind = FALSE)

res_df <- as.data.frame(DESeq2::results(dds))

# Clean Ensembl IDs
res_df$gene_id <- gsub("\\..*", "", rownames(res_df))

# Map gene symbols
res_df$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys      = res_df$gene_id,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

res_df_plot <- res_df %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "Up in Good",
      padj < 0.05 & log2FoldChange < -1 ~ "Up in Poor",
      TRUE ~ "Not Significant"
    )
  )

ggplot(
  res_df_plot,
  aes(log2FoldChange, -log10(padj), color = significance)
) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_text_repel(
    data = subset(res_df_plot, significance != "Not Significant"),
    aes(label = gene_symbol),
    size = 3,
    max.overlaps = 12
  ) +
  scale_color_manual(values = c("grey70", "blue", "red")) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: Good vs Poor TMZ Responders",
    x = "log2 Fold Change (Good vs Poor)",
    y = "-log10 Adjusted P-value",
    color = "significance"
  )

top_10_genes <- res_df %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::arrange(padj) %>%
  dplyr::slice(1:10) %>%
  dplyr::pull(gene_id)

# Clean vsd rownames
vsd_ids_clean <- gsub("\\..*", "", rownames(vsd))

# Match top genes to vsd rows
row_idx <- match(top_10_genes, vsd_ids_clean)
valid <- !is.na(row_idx)

row_idx <- row_idx[valid]
top_10_genes <- top_10_genes[valid]

# Extract expression matrix
heatmap_data <- assay(vsd)[row_idx, ]

# Assign gene symbols as rownames
rownames(heatmap_data) <- res_df$gene_symbol[
  match(top_10_genes, res_df$gene_id)
]

col_annotation <- data.frame(responder = col_data$responder)
rownames(col_annotation) <- rownames(col_data)

# Order samples by responder
ord <- order(col_annotation$responder)


# 10. Heatmap
pheatmap(
  heatmap_data[, ord],
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  show_colnames = FALSE,
  annotation_col = col_annotation[ord, , drop = FALSE],
  annotation_colors = list(
    responder = c(
      Good_responder = "blue",
      Poor_responder = "red"
    )
  ),
  main = "Top 10 Differentially Expressed Genes (by TMZ Response)"
)

```

